---
title: "JavaScript Charting Libraries"
output: html_notebook
---

This notebook is to explore charting library options for exporting for the web.

Start with some data:

```{r}
# connection for SQL Server
library(RODBC)

# for data manipuation
library(dplyr)

# for outputting tables in this notebook
library(knitr)

# for charts
library(ggplot2)

# for branding
library(grid)
library(CRUKBrand)

# for highcharts (also requires dplyr)
library(highcharter)
library(tidyr)

# get data
channel <- odbcConnect("cStream", uid="cstream_admin@cstream.database.windows.net", pwd=.rs.askForPassword("Enter password:"));

#get everything in fairly raw way - note could have got male/female as separate measures here rather than using dplyr later
incidence <- sqlQuery(channel, "
SELECT 
  cs.LEVEL3_CANCER_SITE_CODE,
  ar.AGE_RANGE,
  ar.AGE_FROM,
  y.YEAR,
  g.GENDER_CODE,
  g.GENDER_DESC,
  SUM(f.INCIDENCE_COUNT) as INCIDENCE_COUNT
FROM 
  [dbo].[FCT_INCIDENCE_2017] f 
  INNER JOIN [dbo].[DIM_CANCER_SITE] cs ON f.[CANCER_SITE_WID] = cs.[CANCER_SITE_WID]
  INNER JOIN [dbo].[DIM_YEAR] y ON f.[INCIDENCE_YEAR_WID] = y.[YEAR_WID]
  INNER JOIN [dbo].[DIM_AGE_RANGE] ar ON f.[AGE_RANGE_WID] = ar.[AGE_RANGE_WID]
  INNER JOIN [dbo].[DIM_GENDER] g ON f.[GENDER_WID] = g.[GENDER_WID]
WHERE
  cs.LEVEL1_CANCER_SITE_CODE = 'C00-C97: All Cancers'
GROUP BY
  cs.LEVEL3_CANCER_SITE_CODE,
  ar.AGE_RANGE,
  ar.AGE_FROM,
  y.YEAR,
  g.GENDER_CODE,
  g.GENDER_DESC
");

close(channel);

```

Create a basic chart
```{r}
l3_code <- 'Bone (C40-C41)'
incidence_year <- 2014
min_incidence_year <- incidence_year - 2

incidence_3yr <- incidence %>% 
  filter(YEAR >= min_incidence_year, YEAR <= incidence_year,LEVEL3_CANCER_SITE_CODE == l3_code) %>%
  mutate(INCIDENCE_COUNT_M = ifelse(GENDER_CODE== 'M', INCIDENCE_COUNT, 0),
         INCIDENCE_COUNT_F = ifelse(GENDER_CODE== 'F', INCIDENCE_COUNT, 0)) %>%
  group_by(GENDER_DESC, AGE_RANGE, YEAR) %>%
  summarise(INCIDENCE_COUNT=sum(INCIDENCE_COUNT))

incidence_3yr_mean <- incidence_3yr %>%
  group_by(GENDER_DESC, AGE_RANGE) %>%
  summarise(AVG_INCIDENCE_COUNT=round(mean(INCIDENCE_COUNT)))


hchart(agg_gender_age, "column", x = AGE_RANGE, y = AVG_INCIDENCE_COUNT, group = GENDER_DESC) %>% 
  hc_title(text=paste(l3_code, "Average Number of New Cases per Year")) %>%
  hc_subtitle(text=paste("UK", min_incidence_year, "-", incidence_year)) %>%
  hc_yAxis(title=list(text="Average Number of Cases Per Year")) %>%
  hc_xAxis(title=list(text="Age Range"))

```

Add add export option. Note that order of chaining appears to be important (exporting first, then titles etc.)

```{r}

hchart(agg_gender_age, "column", x = AGE_RANGE, y = AVG_INCIDENCE_COUNT, group = GENDER_DESC) %>% 
  hc_exporting(enabled = TRUE, filename=l3_code) %>%
  hc_title(text=paste(l3_code, "Average Number of New Cases per Year")) %>%
  hc_subtitle(text=paste("UK", min_incidence_year, "-", incidence_year)) %>%
  hc_yAxis(title=list(text="Average Number of Cases Per Year")) %>%
  hc_xAxis(title=list(text="Age Range"))


```

Annoyingly, the CSV and Excel options are not available. Options are PNG, JPEG, SVG, and PDF. Export options only work once rendered in a browser. In demo CSV and EXcel are available http://jkunst.com/highcharter/plugins.html#exporting-csv

Documentation doesn't appear to help:
https://www.rdocumentation.org/packages/highcharter/versions/0.4.0/topics/hc_exporting
http://api.highcharts.com/highcharts/exporting

Try something closer to the example (chart doesn't make sense:)

```{r}

# this is a bit annoying - core classes don't seem to have an easy way to group/split series so need to unpivot the data
library (reshape)
agg_gender_age_pvt <- cast(agg_gender_age, AGE_RANGE~GENDER_DESC)

highchart() %>% 
  hc_xAxis(categories = agg_gender_age_pvt$AGE_RANGE, title=list(text="Age Range")) %>%
  hc_yAxis(title=list(text="Average Number of Cases Per Year")) %>%
  hc_add_series(
    type="column", 
    name = "Male", 
    data = agg_gender_age_pvt$Male) %>%
  hc_add_series(
    type="column", 
    name = "Female", 
    data = agg_gender_age_pvt$Female) %>%
  hc_title(text=paste(l3_code, "Average Number of New Cases per Year")) %>%
  hc_subtitle(text=paste("UK", min_incidence_year, "-", incidence_year)) %>%
  hc_exporting(enabled = TRUE, filename=l3_code)


```


Same result - may need to add the additional javascript libraries for the CSV exports (to test).

Now try some basic theming:


```{r}

CRUKTheme <- hc_theme(
 colors = c("#2E008B", "#EC008C", "#00B6ED", "#A7A8AA", "#AB99D1", "#F799D1", "#99E2F8"),
 chart = list(
  backgroundColor = "#FFFFFF"
 ),
 title = list(
   style = list(
     color = '#2E008B'
   )
 ),
 subtitle = list(
   style = list(
     color = '#2E008B'
   )
 ),
 legend = list(
   itemStyle = list(
     color = '#2E008B'
   ),
   itemHoverStyle = list(
     color = '#2E008B'
   )   
  )
)

hc <- highchart() %>% 
  hc_xAxis(categories = agg_gender_age_pvt$AGE_RANGE, title=list(text="Age Range")) %>%
  hc_yAxis(title=list(text="Average Number of Cases Per Year")) %>%
  hc_add_series(
    type="column", 
    name = "Male", 
    data = agg_gender_age_pvt$Male) %>%
  hc_add_series(
    type="column", 
    name = "Female", 
    data = agg_gender_age_pvt$Female) %>%
  hc_title(text=paste(l3_code, "Average Number of New Cases per Year")) %>%
  hc_subtitle(text=paste("UK", min_incidence_year, "-", incidence_year)) %>%
  hc_exporting(enabled = TRUE, filename=l3_code) #%>%

hc %>% hc_add_theme(CRUKTheme)

```


Next try to export widgets https://www.rdocumentation.org/packages/htmlwidgets/versions/0.3.2/topics/saveWidget

```{r}
library(htmlwidgets)
hcbranded <- hc %>% hc_add_theme(CRUKTheme)
saveWidget(hcbranded, file="widget_self_contained.html", selfcontained=TRUE)
saveWidget(hcbranded, file="widget.html", libdir="lib", selfcontained=FALSE)

```

Notes:
1. SaveWidget feels slow, but does seem to get faster on re-runs (about 7 seconds eventually)
2. Self-contained is 1.7MB
3. Non self-contaiend is 4.4KB
4. Non self-contained directory structure contains the various javascript and css files for each library (e.g. highcharts, jquery). They are not specific to the widget (so no need to republish). 
5. References are relative in the html file (e.g. <script src="lib/htmlwidgets-0.8/htmlwidgets.js"></script>)
6. Structure is consistent so could strip out the div and the related json (which has a unique ID such as data-for="htmlwidget-aa953288dad3c45a794a") and embed that in drupal if don't want to down the iframe route
7. Doesn't appear to create a js object that could be referenced elsewhere (e.g. create own buttons for exporting data)





